The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-16 doc/ResponseLetter-R1.tex

- Line 233: f and f’ are called "adult fertilities". However, they correspond to that rate only for certain specific genotypes. Maybe clarify that by calling them "adult maximum fertilities" ?

\begin{quote}
	{\itshape Good point, fixed.}
\end{quote}

- Results: it took me a while to figure out Fig 1 (mainly because I was looking at it in B and W). I am just wondering whether it would be easier to read with only one f per chart. Maybe as a separate chart that illustrates the general shape of the parameter space for polym * extinction for only one C, one h and one f... then followed by these 6 graphs as they are now.

\begin{quote}
	{\itshape We want to thank the reviewer for this suggestion. We now start with a figure that only has one f, and that separately explains what happens to the allele frequencies (Population genetic outcome) and to the population growth rate (Demographic Outcome), see new Figure 1. We hope that this will make it easier to interpret the old figure 1, which is now Figure 2. We have also included a new supplementary figure (CHECK FIG NUMBERING IN SUPPLEMENTS) with detailed heatmaps for each of the parameter combinations presented in the this figure (i.e., one f per panel). However, we felt that this revised figure was to large and complicated to present in the main text (18 vs. 6 panels).}
\end{quote}

### 2021-12-16 doc/SA-Hermaphrodites-wDemography-R1.tex
Edits to introduction, SA
A first pass at editing the introduction to include more specific
references to empirical SA selection literature. Also some edits to
the other passages.

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{pdflscape}

@@

Fitness trade-offs that impose genetic constraints on adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter) arising from genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, SA selection can arise when beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}.

Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. Self-fertilization in hermaphrodites is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014,Glemin2021}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, antagonism between viability and fecundity selection \citep{Glemin2021}, or spatial heterogeneity and complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019, Glemin2021}. Overall, both theoretical predictions and current empirical data suggest that there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}).

To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. \textcolor{blue}{One of the earliest such models \citep{roughgarden1971density} combined a population genetic model with a density-dependent population model with no intraspecific demographic structure. Charlesworth then combined an age-structured population model with population genetics in a series of papers, consolidated in \cite{BCharlesworth1991}. \cite{orive1995senescence} extended this work to study life-history evolution in organisms with both clonal and sexual reproduction (see also \cite{orive2001somatic}).  \cite{ronce2000kin} added spatial structure and stochasticity to age-structured genetic models to model kin selection, a body of work consolidated in \cite{rousset2004genetic}. Of course there are many other approaches to studying eco-evolutionary dynamics, e.g. combining quantitative genetics models with demographic structure (e.g. \cite{coulson2006putting,barfield2011evolution,childs2016evolution,orive2017effects}) or adaptive dynamics approaches (e.g. \cite{dieckmann2002adaptive,durinx2008adaptive,ten2020habitat}). For a comprehensive review of the eco-evolutionary literature, see \cite{fussmann2007eco}, or more recently, \cite{govaert2019eco}.}

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

In this paper, we extend the model of \citep{deVriesCaswell2019a,deVriesCaswell2019b} to include both dioecious and hermaphroditic species in order to study the parameter conditions under which different outcomes of SA selection, including fixation of male- or female-beneficial alleles as well as balanced polymorphisms, are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 1
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{didacticFig}
 \caption{\footnotesize{Illustration of the possible population genetic outcomes of the model on the left (fixation of either allele, or a polymorphism), and of our demographic outcome of interest on the right:  population growth rates.}}
 \label{fig:didactic}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

@@

% Figure 4
\begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/mimulusinv6FigNewSel.pdf}
 \caption{\footnotesize{Illustration of model predictions using empirically estimated demographic rates for {\itshape M.~guttatus}. Results are shown for two hypothetical populations using demographic rates for locally adapted (Eagle Meadows; panel A), and non-local (Low-Elevation Perennial; panel B) populations reported by \citet{PetersonEtAl2016}. Invasion conditions and extinction thresholds were calculated using the conservative assumption of additive fitness effects in both sexes ($h_f = h_m = 0.5$) for two parameter conditions: obligate outcrossing (solid black lines, dark red shaded region)  and partial selfing with inbreeding depression using the highest field-estimate of selfing, and inbreeding depression parameters ($\delta_i$) calculated for the Iron Mountain population of \citet{Willis1993} (dashed lines, light red shaded region). The location of inv6 is also shown on both plots, using selection coefficients calculated from field estimates of male and female fitness components from \citet{LeeKelly2015} under the relatively conservative assumption of additive fitness effects in both sexes, where $(s_f,\,s_m) = (0.30,\, 0.31)$. Note that in panel A the invasion conditions for partial selfing with inbreeding depression (dotted lines) are nearly indistinguishable from those for obligate outcrossing (solid lines).}} 
 \label{fig:mimulusFig}
 \end{figure}

@@

Despite the demographic pitfalls associated with SA alleles, our example using {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space. The example also appears to provide some empirical support for the conjecure that locally-adapted populations are more likely to harbor SA polymorphisms than marginal or non-locally adapted ones; inviable polymorphic parameter space only occured when using the demographic data for non-local high-elevation perennial populations. Although we cannot make concrete predictions for inv6 in the Iron Mountain population in which it was observed, it is interesting that the estimated SA fitness effects place this polymophic inversion squarely in demographically viable polymorphic parameter space predicted by our model. The available data do not allow for confident estimation of selection coefficients for inv6 (and even require making assumptions about dominance), yet our theoretical predictions are encouragingly consistent with the available data that inv6 is segregating at intermediate frequencies in the large and locally adapted Iron Mountain population.

@@

By combining the tools of demography and population genetics, the framework we present here enables the exploration of interactions between life cycle complexity, mating system, and sexual antagonism. For the sake of simplicity, we made a variety of simplifying assumptions in our analyses, and a variety of extensions are possible. For example, we used a simple life cycle with just two stages for most of this paper, adults and juveniles (the \textit{Mimulus} example has $3$ stages). However, it is possible to include additional age classes, allowing us to explore whether the scope for SA selection to maintain polymorphisms is affected by whether a species exhibits positive or negative senescence \citep{jones2014diversity}. Put another way, does the shape of the survival curve affect the demographic and population genetic consequences of SA selection, and how does this interact with the age at which a sexually antagonistic allele is expressed? What if the allele not only affects male and female fertility but also affects survival of individuals? Additionally, how would the scope for SA polymorphism be affected by density-dependence acting on different life history stages?

>>>>

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{pdflscape}
\usepackage{textgreek}

@@

\textcolor{blue}{Genetic constraints on adaptation arising from conflicting selection} have particularly interesting evolutionary consequences. On the one hand, they provide an effective mechanism for the maintenance of genetic variation, which can increase a population's capacity for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. On the other hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase a population's overall extinction risk \citep{kokko2003sexy,harts2014demography}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Sexually antagonistic selection (abbreviated SA hereafter), \textcolor{blue}{is an important form of conflicting selection that arises when beneficial alleles for one sex are deleterious when expressed in the other. SA selection appears to be} a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and sex-related traits \citep{Kidwell1977, Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, ConnallonClark2012, BondurianskyChenoweth2009, Olito2019}. \textcolor{blue}{Sex-specific estimates of phenotypic selection indicate that sex-differences in directional selection are common \citep[e.g.,][]{DeLisle-etal-2018,Lewis-etal-2011,CoxCalsbeek2009,SinghPunzalan2018}, implying that many genetic variants underlying quantitative traits have SA fitness effects. Similarly, quantitative genetic studies frequently find negative cross-sex genetic correlations for fitness, indicating SA fitness effects for multilocus genotypes (\citealt{Chippendale-etal-2001,Delph-etal-2011}; reviewed in \citealt{ConnallonMatthews2019}). Moreover, polygenic signals of SA selection are evident in recent population genetic analyses of large genomic datasets \citep{RuzickaESEB2020, RuzickaConnallon2022}.}

Alleles with opposing fitness effects through male and female sex functions can also cause SA selection in hermaphrodite populations, where individuals' overall fitness is determined jointy by maternal and paternal reproductive success \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. \textcolor{blue}{In obligately outcrossing hermaphrodite populations, SA selection operates similarly as in seperate-sexed ones (dioecious species). However, the possibility of self-fertilization has important consequences for the maintenance of SA genetic variation in hermaphrodites}. Self-fertilization reduces \textcolor{blue}{opportunities for} balanced SA polymorphisms, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014,Glemin2021}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, antagonism between viability and fecundity selection \citep{Glemin2021}, or spatial heterogeneity and complexity of the life-cycle \textcolor{blue}{can all facilitate the maintenance of} SA polymorphisms \citep{Olito-etal-2018,ConnallonSharmaOlito2019, Glemin2021}. Overall, both theoretical predictions and current empirical data suggest that there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, DelphHerlihy2012, WangBarrett2020, RuzickaESEB2020, RuzickaConnallon2022}.

\textcolor{blue}{While significant efforts have been made to predict when SA polymophisms can be maintained and to detect SA genetic variation from empirical data, much less consideration has been given to the demographic consequences of the maladaptation caused by SA selection}. Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals \textcolor{blue}{which link genotype distributions to population dynamics (e.g., \citealt {johnston2013life, merot2020balancing}). Over the past 50 years, a rich field of eco-evolutionary dynamics has developed which seeks to jointly model the genetic and demographic processes involved in evolutionary change. In particular, theoretical models combining population genetics with density-dependent population growth \citep{roughgarden1971density} and stage-structured populations (consolidated in \citealt{Charlesworth_1994}) paved the way toward understanding the demographic changes accompanying allele frequency changes due to selection. Extensions of these theories to study life-history evolution \citep{orive1995senescence, orive2001somatic}, spatial structure \citep{ronce2000kin}, kin selection \citep{rousset2004genetic}, and quantitative traits \citep[e.g.,][]{coulson2006putting, barfield2011evolution, childs2016evolution, orive2017effects} all highlight the often counterintuitive demographic consequences of adaptive evolution that are orten missed by classical population genetics models (comprehensively reviewed by \citealt{fussmann2007eco} and \citealt{govaert2019eco}).}

%\textcolor{blue}{One of the earliest such models \citep{roughgarden1971density} combined a population genetic model with a density-dependent population model with no intraspecific demographic structure. Charlesworth then combined an age-structured population model with population genetics in a series of papers, consolidated in \cite{Charlesworth_1994}. \cite{orive1995senescence} extended this work to study life-history evolution in organisms with both clonal and sexual reproduction (see also \cite{orive2001somatic}).  \cite{ronce2000kin} added spatial structure and stochasticity to age-structured genetic models to model kin selection, a body of work consolidated in \cite{rousset2004genetic}. Of course there are many other approaches to studying eco-evolutionary dynamics, e.g. combining quantitative genetics models with demographic structure (e.g. \cite{coulson2006putting,barfield2011evolution,childs2016evolution,orive2017effects}) or adaptive dynamics approaches (e.g. \cite{dieckmann2002adaptive,durinx2008adaptive,ten2020habitat}). For a comprehensive review of the eco-evolutionary literature, see \cite{fussmann2007eco}, or more recently, \cite{govaert2019eco}.}

The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few studies since have explicitly incorporated demography into models of sexual antagonism (with the notable exception of \citealt{harts2014demography}). \citet{deVriesCaswell2019a} introduced a mendelian matrix model with intralocus sexual antagonism and population dynamics, but did not analyze the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism. In this paper, we extend the model of \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b} to include both dioecious and hermaphroditic \textcolor{blue}{populations} in order to study the parameter conditions under which different outcomes of SA selection, including fixation of male- or female-beneficial alleles as well as balanced polymorphisms, are also demographically viable. A major strength of our approach is that the model can be parameterized using empirically estimated demographic rates, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata}).

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 1
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{didacticFig}
 \caption{\footnotesize{Illustration of the possible population genetic outcomes of the model on the left (fixation of either allele, or a polymorphism), and of our demographic outcome of interest on the right:  population growth rates. For clarity, results are shown for the case of a dominance reversal, where $h_f = h_m = 1/4$.}}
 \label{fig:didactic}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

@@

% Figure 5
\begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/MimulusInv6LambdaFig.pdf}
 \caption{\footnotesize{Illustration of model predictions using empirically estimated demographic rates for {\itshape M.~guttatus}. Results are shown for two hypothetical populations using demographic rates for locally adapted (Eagle Meadows; panel A), and non-local (Low-Elevation Perennial; panel B) populations reported by \citet{PetersonEtAl2016}. Invasion conditions and extinction thresholds were calculated using the conservative assumption of additive fitness effects in both sexes ($h_f = h_m = 0.5$) for two parameter conditions: obligate outcrossing (solid black lines, dark red shaded region)  and partial selfing with inbreeding depression using the highest field-estimate of selfing, and inbreeding depression parameters ($\delta_i$) calculated for the Iron Mountain population of \citet{Willis1993} (dashed lines, light red shaded region). The location of inv6 is also shown on both plots, using selection coefficients calculated from field estimates of male and female fitness components from \citet{LeeKelly2015} under the relatively conservative assumption of additive fitness effects in both sexes, where $(s_f,\,s_m) = (0.30,\, 0.31)$. Note that in panel A the invasion conditions for partial selfing with inbreeding depression (dotted lines) are nearly indistinguishable from those for obligate outcrossing (solid lines).}} 
 \label{fig:mimulusFig}
 \end{figure}

@@

Despite the demographic pitfalls associated with SA alleles, our example using {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space. The example also appears to provide some empirical support for the conjecure that locally-adapted populations are more likely to harbor SA polymorphisms than marginal or non-locally adapted ones; inviable polymorphic parameter space only occured when using the demographic data for non-local low-elevation perennial populations. Although we cannot make concrete predictions for inv6 in the Iron Mountain population in which it was observed, it is interesting that the estimated SA fitness effects place this polymophic inversion squarely in demographically viable polymorphic parameter space predicted by our model. The available data do not allow for confident estimation of selection coefficients for inv6 (and even require making assumptions about dominance), yet our theoretical predictions are encouragingly consistent with the available data that inv6 is segregating at intermediate frequencies in the large and locally adapted Iron Mountain population.

@@

By combining the tools of demography and population genetics, the framework we present here enables the exploration of interactions between life cycle complexity, mating system, and sexual antagonism. For the sake of simplicity, we made a variety of simplifying assumptions in our analyses, and \textcolor{blue}{many extensions of the model are} possible. For example, we used a simple life cycle with just two stages for most of this paper, adults and juveniles (the \textit{Mimulus} example has $3$ stages). However, it is possible to include additional age classes, allowing us to explore whether the scope for SA selection to maintain polymorphisms is affected by whether a species exhibits positive or negative senescence \citep{jones2014diversity}. Put another way, does the shape of the survival curve affect the demographic and population genetic consequences of SA selection, and how does this interact with the age at which a sexually antagonistic allele is expressed? What if the allele not only affects male and female fertility but also affects survival of individuals? Additionally, how would the scope for SA polymorphism be affected by density-dependence acting on different life history stages?
